Last glacial cycle glacier erosion potential in the Alps

### 2021-05-16 articles/erosion/response-referee2.md

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Ian Evans

Dear Ian Evans,

Thank you very much for your detailed review of our manuscript.

## General Comments

> The authors provide large amounts of very useful results from simulations,
> concerning the effects of varying glacier geometry on erosion potential. It
> was quite an achievement to compute this at fairly high temporal and spatial
> resolution (1 km and 50 a throughout the Alps). The conclusions are
> reasonable, and I value the results. They provide support for the
> traditional view, that high-altitude features such as cirques are eroded by
> steep, local glaciers and not under extensive icefields well above ELA. Most
> erosion is expected along major valleys (glacial troughs): the potential
> erosion is greatest low down the glaciers, with a lesser maximum at high
> altitude and a minimum between. The paper is well written and the Figures
> are all relevant and interesting. The qualitative conclusions are well
> justified.

> An interesting result, not commented on, is that from Figs. 3 and 7 i & j,
> potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm
> sea level equivalent. I would love to see a map of the glaciers at that
> stage\*, which is well below maximum but considerably more than recent ice
> cover (0.3 ± 0.1 mm s.l.e., Farinotti et al. 2019). This peak applies to the
> ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws
> which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).

> As the comment (lines 221-222) on erosion distribution during advance,
> retreat and maxima is of great interest, it would be useful to illustrate
> this with three maps of potential erosion during each of these types of
> phase.

> I find no mention of lithology, rock type, geology, resistance or erodibility
> (excluding that citing Herman on line 56). This omission of half the erosion
> equation should be mentioned; it implies that the ‘potential’ nature of the
> erosion should be emphasised repeatedly. This adds to the admitted
> limitations of lack of feedback to changing topography, absence of hydrology
> and absence of subglacial sediment accumulation or movement. Using constant
> topography might be OK for the last glaciation, as here, but would become
> increasingly unrealistic for several glaciations. Destinations of eroded
> material are not considered: deposition in temporary storage may inhibit
> erosion in parts. The omission of hydrological processes must be a major
> drawback, as Herman et al. (2011), Egholm et al. (2012a, b) and Pedersen et
> al. (2014) showed that its inclusion of subglacial meltwater, plus horizontal
> stress gradients, gave more realistic simulations of erosion. For erosion
> rate as well as for cumulative erosion, it is potential erosion that is
> calculated, rather than predicted erosion.

> A further necessary geographical qualification concerns the misfit between
> simulated and actual glacier extent. The latter is well established from
> mapping of moraines. It is necessary to check Seguinot et al. (2018) to see
> that all the models predict too much ice in the southwest (south of 45° N)
> and far too much in the east (Sava, Drava, Mur, Enns and Traun valleys),
> while the EPICA models produce too little in the west (Rhône and Jura). I
> agree that this is probably because while the amount of precipitation is
> changed, the spatial pattern is not: thus the orographic effect of ice
> build-up increasing precipitation shadow, and the circulation changes due to
> southward shift of westerlies at LGM, are not included. To make this paper
> more self-contained, a brief mention of the 2018 evaluation is in order.
> Once a more realistic LGM precipitation pattern is accepted, future modelling
> might use gradations between that and the present pattern as a function of
> temperature.

> Overall, the paper reports on a very worthwhile computer-intensive exercise.
> It is densely packed with results and repays a careful read and a study of
> the videos. Clearly there can be interesting comparisons with Lai and Anders
> (2021) very recent submission to esurf !

>   [\*That stage can be approximated by pausing the first erosion
>    rates video around 99 ka and 13 ka. The three videos are strongly
>    recommended!]

> The text is short and pithy, so there is space for some extension. There are
> a few missed opportunities, and some incompleteness in the presentation. In
> detail I suggest the following improvements:

## Specific comments - Text:

> The 1 km resolution of the modelling is clearly important, so it is strange
> that the first mention is on p.11. Although it is very demanding of computer
> time, it is still (as admitted) not adequate to represent cirques (averaging
> of the order of 700 m across). Even stranger, there is no mention of the DEM
> used (it was SRTM in the 2018 paper). This information should be provided in
> section 2, p. 3-4.

> The lateral constraints on valley glaciers are usually covered by a ‘form
> factor’: how are they handled by this version of the PISM ice sheet model –
> how has lateral drag been included?

> It might be pointed out that the ‘Koppes’ relationship with an exponent of
> 2.34 is based on 13 data points that are very poorly distributed: in fact,
> Koppes et al. (2015) go on to drop two outliers and reach an exponent of
> 2.62.

> In lines 206-209, the point about trimlines, relating them to
> time-transgressive erosion, is well taken. But they represent the integrated
> effect of many phases of glacial erosion (as noted on 223-224), not
> necessarily just from recessional phases.

## - Figures:

> In Fig. 3 is almost impossible to follow the brown lines, unless advancing
> and retreating parts of lines are distinguished (e.g. by colour). (The ice
> volume video may help out.)  Also: keep to ‘potential erosion volume’. The
> pale brown bars show a decline in annual erosion volume with ice volumes over
> 1.8 cm s.l.e., stronger than is implied in line 110 of the text – it is more
> than ‘slight’.

> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic
> map or hillshade of the Rhine area on the same scale should be provided.

> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
> unhealthy convex glacier. Fascinating, but surely a more representative
> cirque should be used. Actually these three photos are not well integrated
> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
> (vertical-sided) trough example, over-used by textbooks.

> Fig. 5: The elevation histogram in Fig. 5a could be misleading, as it seems
> to cover the whole study area, including parts never glacier-covered. It
> should be limited to areas glacier-covered at the maximum, and preferably
> supplemented by a histogram for those covered now or at the minimum. The
> accompanying main graph (5a) relates to areas glacier-covered at each time
> slice. Re another reviewer’s comment, yes LGM glaciers did reach below 500 m:
> at L. Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below
> 100 m. An integrated plot of total-cycle potential erosion rate (and
> volume?) at each altitude, over glacier-covered areas, would be interesting
> (as an extra Figure): can rates at high altitude, where glacial duration is
> longer, catch up with those at low altitude?  Probably not, given the
> concentration of flow down major valleys. Striping above 3000 m, presumably
> related to low surface area, suggests a switch there to 20 m or 40 m bands
> would be advisable. Caption; ‘modelled potential erosion rates’?

> In Fig. 7, the quantitative contrasts between the ‘Koppes’ law (e) and the
> others are alarming. If the scales are taken literally, (e) gives potential
> erosion around 1 m, while (f) and (g) give hundreds of m, in 120 ka. As 120
> ka is only a fraction of the Quaternary, hundreds of m seems too much, while
> 1 m seems very low (0.008 mm a-1 overall). For me, there should be a correct
> answer between these extremes. The spatial pattern, however, is more
> important.

> Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
> As there are mountains at 3630, 3583 and 3192 m around the sources, it is
> likely that   there would be steeper ice slopes up to these peaks, giving
> rather different stresses and erosion potentials. Also are the instabilities
> in Fig. 8b, near source and terminus, edge effects or related to the 1 km
> resolution?

>   The three videos are well worth watching  -in fact they clarify some
>   queries arising from the paper. Perhaps they could be captioned in the
>   ‘Video supplement’ paragraph. Some extra information on the third (bedrock
>   altitude) video could point out that there are numerous zero values not
>   shown: the zero geometric mean rates from 2300 to 3000 m around 24 ka
>   otherwise seem to conflict with the numerous positive rates plotted. Zero
>   rates presumably show where ice is frozen to the bed: their altitudinal
>   distribution might be worth discussion in the text. Above 3000 m means are
>   positive, which seems strange.

## Technical corrections:

> All map figures need km scale bars.

> Line 24  ‘are yet’

> 27 ‘in the absence’

> 38 ‘Examination … suggests’

> 61 ‘modelled erosion potential’ rather than ‘ modelled erosion rates’ ?

> 70 insert ‘isostatic’?

> 72 ‘and is run to…’

> 78 ‘temperature lapse-rate’ or is it ‘temperature change’ ?

> 116-7 ‘much of the’; and delete comma after bracket.

> 161 ‘does not increase’

> 190 delete both commas

## References not in original:

> Egholm, D.L., Pedersen, V.K., Knudsen, M.F., Larsen, N.K., 2012a. Coupling
> the flow of ice, water, and sediment in a glacial landscape evolution model.
> Geomorphology 141–142: 47–66.
> http://dx.doi.org/10.1016/j.geomorph.2011.12.019.

> Egholm, D.L., Pedersen, V.K., Knudsen, M.F., Larsen, N.K., 2012b. On the
> importance of higher-order ice dynamics for glacial landscape evolution.
> Geomorphology 141–142: 67-80.

> Farinotti, D., et al., 2019. A consensus estimate for the ice thickness
> distribution of all glaciers on Earth. Nat. Geosci. 12, 168–173.

> Lai, J., Anders, A.M., 2021. Climatic controls on mountain glacier basal
> thermal regimes dictate spatial patterns of glacial erosion.
> https://esurf.copernicus.org/preprints/esurf-2021-26/

> Pedersen, V.K., Huismans, R.S., Herman, F., Egholm, D.L., 2014. Controls of
> initial topography on temporal and spatial patterns of glacial erosion.
> Geomorphology 223, 96-116.

## Acknowledgements:

> I am grateful for helpful discussions with Iestyn Barr, Jeremy Ely, Matt
> Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek.

> - Ian S. Evans, Durham University, U.K.

### 2021-05-17 articles/erosion/response-referee1.md

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

## Specific comments

> Line 24 Maybe a bit odd to use 4 very old references from the Alps only (you
> hardly used more than 2 references throughout the MS and there are also high
> variety of glacial landforms in other mt ranges as you mentioned).

We agree the references are not directly relevant. The citation was reduced to
the latest reference by Penck (1905) and moved within the sentence to clarify
that it is specific to the Alps.

@@

> Line 33 I think that’s not very easy to understand what landform you refer
> to? The “periglacial blockfields topped by glacial erratic boulders” (Wirsig
> et al., 2018)? Eventually be more precise on this would help readers.

Agreed. "Characteristic landform preservation" was replaced with "preserved
periglacial blockfields".

@@

> Line 101 Could be misleading as (glacial and periglacial) cirque erosion
> processes are not really covered by any glacial erosion law discussed here
> (or any other I am aware of, Sanders et al., 2012).

Thank you for pointing out this study. The lack of cirque erosion processes was
acknowledged and the reference added in this sentence, as well as in the
relevant part of the discussion on the age of the glacial landscape.

@@

> Line 127 Higher precipitation increases ice flux and thus erosion, I guess?
> Would eventually helpful to mention (even though it might be referred in
> Seguinot et al., 2018).

Indeed. The increase in ice discharge was made explicit.

> Line 140 Can you be more precise what you mean by realistic? You mean because
> of the localized pattern? Maybe I am wrong but shouldn’t the erosion
> potential at least at the lake Constance overdeepening (Fig. 7) not in the
> order of hundreds of meters (cf. Preusser et al., 2010) and the best fit
> rather (b) or (c) – at least from what I read from the transect (e-h)?
> Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also
> presented in meters (annotation at the bar is missing). For example, in (b)
> the maximum erosion potential is like >>1000 m (if in meters) while in
> corresponding (f) it seems clearly lower than 1000 m.

TODO: check the numbers here. Make sure that the maps (a-d) fit the profiles
(e-h), and then think again how to address the comment.

@@

> Furthermore, isn’t the impression of the fit potentially very dependent on
> the initial model parametrization (Seguinot et al., 2018), i.e. the ice flux
> velocity? You should address these dependences!

TODO: discuss uncertainty with sliding.

> Line 163 Observing Alpine topography I find this result important, which can
> maybe also serve as explanation why e.g. (low erodible) areas away from the
> big troughs covered during glacial maxima only, do surprisingly often show
> no/very low degree in glacial modification (e.g. Ticino; Kelly et al., 2004).

The following sentence was addedin this paragraph. "Nevertheless, it may
explain why some areas covered during glacial maxima only, such as some valleys
on the southern side of the Alps, appear to have experienced only little
glacial modification of their topography."

@@

> Line 187 “time-transgressive radial pattern”. I don’t understand what you
> mean..

This sentence was only meant to recall the aforementioned results. We rephrased
to "the erosion patterns shift spatially through the glacial cycle".

@@

> Line 195 Very much share this view!

Good to know!

> Line 197 Would recommend to be more precise. Guess you know that there are
> many, many cirques in the Alps as low as 1500 m (and even below) e.g. forming
> in areas outside the connected ice stream network in the SW and easternmost
> Alps. This can be revealed by a quick check at any higher resolution DEM or
> google earth.

The statement was reworked and restricted to "the higest mountain cirques".

@@

> In the MS the word “yet” is very often used. Eventually consider reducing.
> The frequency is a bit irritating when reading.

In several instances we replaced "yet" with "but" and "however".

@@

## Figures

> Fig. 2 Please indicate the outline of the connected ice stream network during
> the LGM (like in Fig. 4 of Seguinot et al., 2018). Especially in the eastern
> and SW Alps coverage largely deviates from what has been suggested from field
> data compilation (e.g. Ehlers and Gibbard, 2004). Even though outcomes might
> not be changing much, it might be helpful to know how much the %overlap is -
> erosional potential might change as fluvial topography turns into glacial one
> (e.g. Harbor, 1988) and this is probably not what you want to mix, I guess.

We added the LGM outline.

TODO: I don't understand the second part of the comment.

@@

> Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as
> low as 500 m.a.s.l (and even below) from 110 – 40 ka in the Seguinot et al.,
> 2018 model?

Yes. This primarily concerns the southern slope of the Alps where topography
drops abruptedly from the highest peaks to deep valleys. It concerns only a few
grid cells for a given time step and is certainly not a realistic model output.

TODO: test cutting the hypsogram below a minimum number of grid cells.
